---
title: "1.1 Preliminaries and Software installation"
author: Adapted from Software Carpentry; modified by Maina,Stephanie
date: "27 October 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**How this is going to go down**

We are going to cover the material in a different way than you would get taught a statistics course, and a different way than you would get taught a programming course.

We're going to start pretty slow, and slowly ramp the pace up. The idea is to get you to a point where you can continue learning yourself. It is not possible to learn to program in three days. It will take time and practice. 

If you get lost in the material at the second half of the day, don't despair, make sure you speak up.

We're going to focus on concepts, best practices and workflow almost as much on getting you going with the syntax and commands. R is simply too large to teach in three days that the full course runs for. Our main hope is that you get enough of a flavour of it to continue the learning process yourself. We probably have more material here than we'll get through today, and we feel that this really just scratches the surface.

**Learning R can be frustrating**

Learning R is not necessarily hard, but not necessarily easy either. Different people will make the logical connections faster than others, and until it "clicks" it may seem like a battle.
Ask questions as you are unclear and we'll help make those connections faster.
It is a programming language
don't think of it a statistical program that you use from a command line
think of it as a programming language that happens to have a lot of statistical functions.
The big challenge is going to be bridging between the nebuluous "what I think I want to do" and the precice "computer - this is what I need you to do".
It is easy to underestimate how precice instructions need to be. We'll see examples, but things like ("take the mean of the leg lengths by species" are intuitively obvious to us, but can be hard precicely convey to a computer.
what if there are missing data?
does it matter if there are different numbers individual per species?
does it matter if there are different sized legs on each species (which leg?)
what happens if someone gave ranges for some of the leg measurements, rather than one number?
what about subspecies?
There is a good Douglas Adams quote about this: "If you really want to understand something, the best way is to try and explain it to someone else. That forces you to sort it out in your mind. And the more slow and dim-witted your pupil, the more you have to break things down into more and more simple ideas. And that's really the essence of programming. By the time you've sorted out a complicated idea into little steps that even a stupid machine can deal with, you've learned something about it yourself." (from Dirk Gentley's Holistic Detective Agency).

**INSTALLING R and R Studio**

R and R Studio are separate packages. You will need to install R first.
R is the basic package we are using. R Studio is an add-on that make R easier to use for
beginners. 
These instructions should work for Windows and MAC users for installing R and R Studio.

#### **1.INSTALLING R:**

  + Go to http://www.r-project.org/
    - on the "Getting Started" box, click on "download R." 
    - on https://cran.r-project.org/mirrors.html, find a site of your choice, ideally the closest to your actual location.
    - Click to go to that site
    
  + On the mirror site of your choice, click on your operating system:
    - Download R for Linux
    - Download R for (Mac) OS X
    - Download R for Windows
    
    

+ If **Windows**:

    - click on "**base**" and then on **Download R 3.4.1 for Windows**. 
    - save the file "**R-3.4.1-win.exe**" and then click on it to execute it. 
    - Once the dialog box opens, click "**RUN**." A Setup Wizard should appear. 
    - Keep clicking "**Next**" (or change features if you understand them), until it is finished.
    
        **Note** that 3.4.1 is the current version as this is being written, but use whatever shows up as current.
        
+ If **Mac**:

    - click on the last version of the R package **R-3.4.1.pkg** on the page **R for Mac OS X** under the **Files** section
    - Save the **.pkg file**
    - Double-click it to open, and follow the installation instructions.
    
Now that **R** is installed, you should now see an icon on your desktop, with a large capital "**R**".


You then need to download and install RStudio.
  

#### **2. INSTALLING R STUDIO:**

We're going to focus on using RStudio for two reasons:

It makes my life easier: you are all using the same tools and we'll be able to help you with your problems faster.

It will make your life easier: it's got lots of features that help people, especially beginners. It will help you organise your work, develop good workflows. On the other hand, it's not very intrusive and if you use a different interface (such as the plain R interface that you installed) it will feel very similar.

  + Go to http://www.rstudio.com
  + click on "Download RStudio" and follow the directions for your operating system.


**USING R AND R STUDIO**
Open R Studio by clicking on the icon.
You're ready to go!

**Getting started with RStudio**

Load R studio however you do that on your platform.

RStudio has four panes:

  + 1. Bottom left - this is the R interpreter. If you type code here, it is "evaulated" so that you get an answer.
  + 2. Top left - editor for writing longer pieces of code.
  + 3. Top right will tell you things about objects in the workspace. We'll get to this soon, but this will be things like data objects, or functions that will process them. It is completely unrelated to the file system. The "History" tab will keep an eye on what you've done.
  + 4. Will display files, plots, packages, and help information, usually as needed.
We'll do everything in a project, as that will help when we get some data.

  + "Project": "Create Project."
  + choose "New Project, (start a project in a new directory)".
  + Leave the "Type" as the default.
  + In the "Directory name" type the name for the project (in our case `g2g` might be a good name).
  + In the "Create project as a subdirectory of" field select (type or browse) for the parent directory of the project. By default this is probably your home directory, but you might prefer your Documents folder.
The RStudio window morphs around a bit, and the top left pane will disappear.

In the bottom right panel, make sure that the "Files" tab is selected and make a new folder called `data`. I strongly recommend keeping a directory like this in every project that contains your data. Treat it read only (that is, write once, then basically don't change). This may be a large shift in thinking if you've come from doing data analysis and management in Excel.

In more complicated projects, I would generally have a folder called `R` that contains scripts and function definitions, another called `figs` that contained figures that I've generated, and one called `doc` that contains the manuscript, talks, etc.

We're going to spend a bit of time using a data set. You can download this from here and put it into that directory. Download this file, open the `g2g/data` folder and move it there (if you click More: show folder in new window, you'll get a file browser window opening in about the right place). Similarly, also grab [this file](https://nicercode.github.io/intro/data/seed_root_herbivores.txt) and put it in the `data` directory too.

To make sure everything is working properly, in the console window type.

```{r results='hide'}
file.exists("data/seed_root_herbivores.csv")
```

which should print

```{r echo=F}
file.exists("data/seed_root_herbivores.csv")
```



#### 3. Installing and managing packages

***Ex01_0***

  + **LISTS OF PACKAGES**
  
```{r eval=FALSE}
browseURL("http://cran.r-project.org/web/views/")  # Opens CRAN Task Views in browser
# Opens "Available CRAN Packages By Name" (from UCLA mirror) in browser
browseURL("http://cran.stat.ucla.edu/web/packages/available_packages_by_name.html")
# See also CRANtastic
browseURL("http://crantastic.org/")
```

  + **See current packages**
```{r eval=FALSE}
library()  # Brings up editor list of installed packages
search()   # Shows packages that are currently loaded
```


  + **TO INSTALL AND USE PACKAGES**
  
```{r eval=FALSE}
# Can use menus: Tools > Install Packages... (or use Package window)
# Or can use scripts, which can be saved in incorporated in source
install.packages("ggplot2")  # Downloads package from CRAN and installs in R
?install.packages
library("ggplot2")  # Make package available; often used for loading in scripts
require("ggplot2")  # Preferred for loading in functions; maybe better?
library(help = "ggplot2")  # Brings up documentation in editor window
```

  + **UPDATE PACKAGES**
  
```{r eval=FALSE}
# In RStudio, Tools > Check for Package Updates
update.packages()  # Checks for updates; do periodically
?update.packages
```

  + **UNLOAD/REMOVE PACKAGES**
  
```{r eval=FALSE}
# By default, all loaded packages are unloaded when R quits
# Can also open Packages window and manually uncheck
# Or can use this code
# To unload packages
detach("package:ggplot2", unload = TRUE)
?detach
```

  + **To permanently remove (delete) package**
  
```{r eval=FALSE}
install.packages("psytabs")  # Adds psytabs
remove.packages("psytabs")   # Deletes it
?remove.packages
```

# 4. Getting help with R
This is a summary of the information you can find here https://www.r-project.org/help.html

#### R Help: `help()` and `?`
The `help()` function and `?` help operator in R provide access to the documentation pages for **R functions, data sets, and other objects**, both for packages in the standard R distribution and for contributed packages. 
For example, to access documentation for the standard mean function, for example, enter the command:`help(mean)` or `help("mean")`, or `?mean` or `?"mean"` (the quotes are optional).

You may also use the `help()` function to access information about a **package** in your library - for example, `help(package="MASS")` - which displays an index of available help pages for the package along with some other information.

Help pages for functions usually include a section with executable examples illustrating how the functions work. You can execute these examples in the current R session via the `example()` 
```{r}
example(mean)
```

#### Searching for Help Within R

The `help()`  function and `?` operator are useful only if you already know the name of the function that you wish to use. There are also facilities in the standard R distribution for discovering functions and other objects. The following functions cast a progressively wider net. Use the help system to obtain complete documentation for these functions: for example, `apropos()`or `?apropos`.

The `apropos()` function searches for objects, including functions, directly accessible in the current R session that have names that include a specified character string. This may be a literal string or a regular expression to be used for pattern-matching. By default, string matching by `apropos()` is case-insensitive. For example, 

```{r}
apropos("^glm") 
```
returns the names of all accessible objects that start with the (case-insensitive) characters "glm"

#### R Help on the internet

There are internet search sites that are specialized for R searches, including `search.r-project.org` (which is the site used by RSiteSearch) and `Rseek.org`.

It is also possible to use a general search site like **Google**, by qualifying the search with "R" or the name of an R package (or both). 
**It can be particularly helpful to paste an error message into a search engine** to find out whether others have solved a problem that you encountered.



# 5. EXITING R AND R STUDIO
You can exit both at the same time using File??? Quit R Studio
Or you can type q() at the command prompt. Note that this is the letter q followed by open and
closed parentheses. 